Boosting Higgs Pair Production in the $b\bar{b}b\bar{b}$ Final State with Multivariate Techniques

### 2015-12-14 paper/sec-conclusions.tex
completed Daniela's comments
more work in polishing text

In this work we have presented a feasibility study for
 the measurement of Higgs pair production in the $b\bar{b}b\bar{b}$
final state at the High-Luminosity LHC with $\mathcal{L}=3$ ab$^{-1}$.
%
Our strategy was based on the combination of a traditional
cut-based analysis with state-of-the-art multivariate techniques.
%
We take into account 
all relevant backgrounds, in particular
the irreducible $4b$
and the reducible 
$2b2j$ and $4j$ QCD multijets.
%
We illustrate how the $2b2j$ component leads to
a contribution comparable to that of the $4b$ process,
due to a combination of  parton shower effects, $b$-quark 
pair radiation, and selection requirements.
%
Since any realistic study of the feasibility of Higgs pair
production at the HL-LHC should take into account the
high-PU environment in which such measurement will be performed,
in this work we have demonstrated the robustness of our analysis strategy
under high-PU conditions.

Combining the contributions from the resolved,
intermediate and boosted categories, we find that, for
$\mathcal{L}=3$ ab$^{-1}$, the
signal significance for
the production of Higgs pairs can be as large as $S/\sqrt{B}\simeq 4.0$.
%
This indicates that already from the $b\bar{b}b\bar{b}$
final state alone
it should be possible to claim evidence for Higgs pair production at
the HL-LHC.
%
Our study also suggests possible avenues that the LHC experiments
could explore to further improve this signal significance.
%
One handle could be to reduce the contribution from light and charm
jet mis-identification, ensuring that the
$2b2j$ component is subdominant over the $4b$ background.
%
Another possibility would be to improve the mass resolution of the Higgs
reconstruction
in high-PU environments, and, more in general,
to optimize the PU subtraction
strategy, in order
to reduce the impact of PU in the modelling
of kinematical variables and the associated
degradation of the MVA discrimination.
%
Following these and related improvements in the experimental
analysis, it might be feasible to achieve
an observation of Higgs pair production even with only
$\mathcal{L}=300$ fb$^{-1}$.

@@

{\bf\noindent  Acknowledgments \\}
We thank F.~Bishara, R.~Contino, A.~Papaefstathiou and
G.~Salam for useful discussions on the topic
of Higgs pair production.
%
We thank E.~Vrodynou and M.~Zaro for
assistance with di-Higgs production
  in {\tt MadGraph5\_aMC@NLO}.
%
  J.~R. is supported by an STFC Rutherford Fellowship and
  Grant ST/K005227/1 and ST/M003787/1.
%
J.~R. and N.~H. are
supported by an European Research Council Starting Grant ``PDF4BSM".
%
K.~B. is supported by a Rhodes Scholarship.
%
J.~F. and C.~I. are supported by the STFC.

>>>>

In this work we have presented a feasibility study for
 the measurement of Higgs pair production in the $b\bar{b}b\bar{b}$
final state at the High-Luminosity LHC with $\mathcal{L}=3$ ab$^{-1}$.
%
Our strategy is based on the combination of traditional
cut-based analysis with state-of-the-art multivariate techniques.
%
We take into account 
all relevant backgrounds, in particular
the irreducible $4b$
and the reducible 
$2b2j$ and $4j$ QCD multijets.
%
We have illustrated how the $2b2j$ component leads to
a contribution comparable to that of the $4b$ process,
due to a combination of  parton shower effects, $b$-quark 
pair radiation, and selection requirements.
%
We have also demonstrated the robustness of our analysis strategy
under high-PU conditions.

Combining the contributions from the resolved,
intermediate and boosted categories, we find that, for
$\mathcal{L}=3$ ab$^{-1}$, the
signal significance for
the production of Higgs pairs can be as large as $S/\sqrt{B}\simeq 4.0$.
%
This indicates that, already from the $b\bar{b}b\bar{b}$
final state alone,
it should be possible to claim evidence for Higgs pair production at
the HL-LHC.
%
Our study also suggests possible avenues that the LHC experiments
could explore to further improve this signal significance.
%
One handle would be to reduce the contribution from light and charm
jet mis-identification, ensuring that the
$2b2j$ component is subdominant over the $4b$ background.
%
Another possibility would be to improve the mass resolution of the Higgs
reconstruction
in high-PU environments, and, more in general,
to optimize the PU subtraction
strategy, in order
to reduce the impact of PU in the modelling
of kinematical variables and the associated
degradation of the MVA discrimination.
%
Such experimental improvements might
even make possible
an observation of Higgs pair production already at the
end of Run II with
$\mathcal{L}=300$ fb$^{-1}$.

@@

{\bf\noindent  Acknowledgments \\}
We thank F.~Bishara, R.~Contino, A.~Papaefstathiou and
G.~Salam for useful discussions on the topic
of Higgs pair production.
%
We thank E.~Vrodynou and M.~Zaro for
assistance with di-Higgs production
  in {\tt MadGraph5\_aMC@NLO}.
%
  J.~R. is supported by an STFC Rutherford Fellowship and
  Grant ST/K005227/1 and ST/M003787/1.
%
J.~R. and N.~H. are
supported by an European Research Council Starting Grant ``PDF4BSM".
%
K.~B. is supported by a Rhodes Scholarship.
%
D.~B., J.~F. and C.~I. are supported by the STFC.

### 2015-12-14 paper/sec-mva.tex
more work in polishing text

In this section we study how the MVA results are modified when
signal and background events are embedded in a high-PU environment.
%
The simulation of PU and the corresponding subtraction with
{\tt SoftKiller} have been performed as described
in Sect.~\ref{sec:pileup}.
%
The cut-based analysis and the subsequent
MVA optimization have been performed using exactly the same
settings as in the case without PU.
%
In Table~\ref{table:cutflowMVA_PU} we provide the results
  at the end of the cut-based analysis,
  for the case with $\la n_{\rm PU}\ra=80$ with SK
  subtraction.
%
It is clear that the pre-MVA 
signal significance has been degraded
as compared to the case without PU.
%
We now find values for $S/\sqrt{B}$ of 0.25, 0.15 and 0.31, in the resolved,
intermediate and boosted categories, respectively, to be compared
with the corresponding values without PU, namely 1.0, 0.9 and 1.0,
see Table~\ref{table:cutflowMVA}. 
%

Subsequently to 
 the cut-based analysis, the signal and background
 MC events in the case of PU satisfying all selection cuts
  are used to re-train the MVA.
%
In Table~\ref{table:cutflowMVA_PU}
we also provide  the number of signal and
    background events expected
    at the HL-LHC with $\mathcal{L}=3$ ab$^{-1}$,
    as well as the
    corresponding values for $S/\sqrt{B}$ and $S/B$,
    for both the cut-based analysis (corresponding
    to $y_{\rm cut}=0$) and after the
    optimal MVA cut.
    %
    From Table~\ref{table:cutflowMVA_PU} we observe that, thanks
to the MVA, the signal significance and the
signal over background ratio is improved from 0.31 and $3\cdot 10^{-4}$
(0.25 and $10^{-4}$) up to 2.9 and 3.4\% (2.5 and 1.5\%)
in the boosted (resolved) category.
%
The intermediate category, on the other hand, exhibits a
significantly lower post-MVA value of $S/\sqrt{B}\simeq
1.1$.
%
We thus conclude that, as was already the case
without PU,
the boosted category is the most promising
one for the study of Higgs pair production in the $b\bar{b}b\bar{b}$
final state
at the HL-LHC, also for the case of a high-PU environment, but that
important information will also be obtained from
the resolved category, which has a comparable significance.

@@

Combining the three categories, resolved,
intermediate and boosted, we determine the overall signal
significance, finding
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0~(1.3) \, ,\quad
\mathcal{L}=3000~(300)\,{\rm fb}^{-1}\, .
\ee
%
We thus conclude that the measurement of
Higgs pair production in the $b\bar{b}b\bar{b}$ final state at the HL-LHC
should be 
well above the threshold for evidence.
%
On the other hand, such measurement seems to be out of reach
for the total integrated luminosity expected by the end of Run II.

In Table~\ref{table:cutflowMVA_fakes} we compare,
after the MVA is applied,
the total
number of background events, $N_{\rm ev}^{\rm tot}$,
    expected at the HL-LHC, with
     $N_{\rm ev}^{\rm 4b}$, the corresponding number
    from the QCD $4b$ component, in the cases with and without
    PU.
    %
    The MVA has been trained to the total background sample,
    though differences
    in the kinematic distributions of the $4b$ and $2b2j$ processes are small,
    see Fig.~\ref{fig:histoBack}.
    %
    The major improvement of eliminating the contamination
    of light and charm jet mis-identification is for the boosted and resolved categories.
%
    In the case of PU+SK, combining the three categories but assuming that
    the only relevant background is the irreducible QCD $4b$ component,
    we obtain
    \be
\lp \frac{S}{\sqrt{B_{\rm 4b}}}\rp_{\rm tot} \simeq 5.3~(1.8) \, ,\quad
\mathcal{L}=3000~(300)\,{\rm fb}^{-1}\, ,
\ee
Therefore, a discovery of Higgs pair production at the
HL-LHC might be possible if the
mis-identification of light and charm jets as
$b$-jets can be  reduced.
%
With these assumptions, and provided
the signal efficiency can be further enhanced,
this measurement might even be feasible
by the end of Run II.
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
  \centering
  \small
  \begin{tabular}{|c|c|c|c||c|c||c|c|}
        \hline
        Category  &    &  $N_{\rm ev}^{\rm tot}$  &  $N_{\rm ev}^{\rm 4b}$   &
        $S/\sqrt{B_{\rm tot}}$ & $S/\sqrt{B_{\rm 4b}}$  
        &  $S/B_{\rm tot}$ & $S/B_{\rm 4b}$\\ 
    \hline
    \hline
    \multirow{2}{*}{Boosted} &  no PU  & $2.2\cdot 10^4$  & $1.5\cdot 10^4$     & 
      5.4 &  6.5 & 0.034 & 0.05 \\
    & PU+SK & 7300  &  500  &  2.9  & 3.6 &  0.034 & 0.04 \\
    \hline
    \hline
    \multirow{2}{*}{Intermediate} &  no PU   & 4500  & 2000    &
    4.8  &  5.3 &  0.06  &  0.08 \\
    & PU+SK  & 2600   &  1500  & 1.1  & 1.2 & 0.02 & 0.03 \\
    \hline
    \hline
    \multirow{2}{*}{Resolved} &   no PU  & $8\cdot 10^4$   &
    $4.5\cdot 10^4$
    & 4.5  & 6.6  & 0.01 & 0.03 \\
    & PU+SK  &  $3\cdot 10^4$   &   $1.3\cdot 10^4$ & 2.5    & 3.7  & 0.015 & 0.03 \\
    \hline
      \end{tabular}
  \caption{\small Post-MVA number of total background events, $N_{\rm ev}^{\rm tot}$,
    expected at the HL-LHC, compared
    with $N_{\rm ev}^{\rm 4b}$,
    the number of QCD $4b$ background events.
    %
    We also indicate the corresponding signal significance and signal over background
    ratios.
        \label{table:cutflowMVA_fakes}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

In this section we study how the MVA results are modified
when the analysis is performed in a high-PU environment.
%
The cut-based analysis and the subsequent
MVA optimization have been performed using the same
settings as in the case without PU.
%
In Table~\ref{table:cutflowMVA_PU} we provide the results
  at the end of the cut-based analysis,
  for the case with $\la n_{\rm PU}\ra=80$ with SK
  subtraction,
  as well as the post-MVA results for the
  corresponding optimal values of $y_{\rm cut}$.
  %
  In addition, we also quote in
  both cases the number of signal and
    background events expected
    at the HL-LHC with $\mathcal{L}=3$ ab$^{-1}$
    and the values of $S/\sqrt{B}$ and $S/B$

We observe how the pre-MVA 
signal significance is degraded
as compared to the case without PU.
%
We now find values for $S/\sqrt{B}$ of 0.25, 0.15 and 0.31, in the resolved,
intermediate and boosted categories, respectively, to be compared
with the corresponding values without PU, namely 1.0, 0.9 and 1.0.
%
Once the MVA is applies, the signal significance in the three
categories is increased to 2.5, 1.1 and 2.9 respectively.
%
We also observe that, thanks
to the MVA, the signal significance and the
signal over background ratio is improved from 0.31 and $3\cdot 10^{-4}$
(0.25 and $10^{-4}$) up to 2.9 and 3.4\% (2.5 and 1.5\%)
in the boosted (resolved) category.
%
The intermediate category, on the other hand, exhibits a
significantly lower post-MVA value of $S/\sqrt{B}\simeq
1.1$.
%
We thus conclude that, as was already the case
without PU,
the boosted category is the most promising
one for the study of Higgs pair production in the $b\bar{b}b\bar{b}$
final state
at the HL-LHC, also for the case of a high-PU environment, but that
important information will also be obtained from
the resolved category, which has a comparable significance.

@@

In Table~\ref{table:cutflowMVA_fakes} we compare,
after the MVA is applied,
the 
number of background events  expected at the HL-LHC,
both the total number, $N_{\rm ev}^{\rm tot}$,
and those from the  QCD $4b$ component
$N_{\rm ev}^{\rm 4b}$, for the analyses without and
with PU.
%
 We also quote the corresponding values of the signal 
    significance and the signal over background ratio.
    %
    Note that the MVA has been trained to the total background sample,
    though differences
    in the kinematic distributions of the $4b$ and $2b2j$ processes are small,
    see Fig.~\ref{fig:histoBack}.
    %
    The major improvement from eliminating the contamination
    from light and charm jet mis-identification arises
    for the boosted and resolved categories.
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{table}[h]
  \centering
  \small
  \begin{tabular}{|c|c|c|c||c|c||c|c|}
        \hline
        Category  &    &  $N_{\rm ev}^{\rm tot}$  &  $N_{\rm ev}^{\rm 4b}$   &
        $S/\sqrt{B_{\rm tot}}$ & $S/\sqrt{B_{\rm 4b}}$  
        &  $S/B_{\rm tot}$ & $S/B_{\rm 4b}$\\ 
    \hline
    \hline
    \multirow{2}{*}{Boosted} &  no PU  & $2.2\cdot 10^4$  & $1.5\cdot 10^4$     & 
      5.4 &  6.5 & 0.034 & 0.05 \\
    & PU+SK & 7300  &  500  &  2.9  & 3.6 &  0.034 & 0.04 \\
    \hline
    \hline
    \multirow{2}{*}{Intermediate} &  no PU   & 4500  & 2000    &
    4.8  &  5.3 &  0.06  &  0.08 \\
    & PU+SK  & 2600   &  1500  & 1.1  & 1.2 & 0.02 & 0.03 \\
    \hline
    \hline
    \multirow{2}{*}{Resolved} &   no PU  & $8\cdot 10^4$   &
    $4.5\cdot 10^4$
    & 4.5  & 6.6  & 0.01 & 0.03 \\
    & PU+SK  &  $3\cdot 10^4$   &   $1.3\cdot 10^4$ & 2.5    & 3.7  & 0.015 & 0.03 \\
    \hline
      \end{tabular}
  \caption{\small Post-MVA number of background events
    expected at the HL-LHC, comparing the total number, $N_{\rm ev}^{\rm tot}$,
    with those from the QCD $4b$ background
    component only, $N_{\rm ev}^{\rm 4b}$.
     %
    We also quote the values of the signal 
    significance and signal over background ratios in each case.
        \label{table:cutflowMVA_fakes}
  }
\end{table}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

Combining the three categories, taking into
account all background components, we obtain the overall signal
significance:
\be
\lp \frac{S}{\sqrt{B}}\rp_{\rm tot} \simeq 4.0~(1.3) \, ,\quad
\mathcal{L}=3000~(300)\,{\rm fb}^{-1}\, ,
\ee
%
indicating  that the measurement of
Higgs pair production in the $b\bar{b}b\bar{b}$ final state at the HL-LHC
should be 
well above the threshold for evidence.
%
On the other hand, such measurement seems to be
extremely challenging
for the total integrated luminosity expected by the end of Run II.
%
Under the  assumption that
    the only relevant background would be the irreducible QCD $4b$ component,
    we obtain instead:
    \be
\lp \frac{S}{\sqrt{B_{\rm 4b}}}\rp_{\rm tot} \simeq 5.3~(1.8) \, ,\quad
\mathcal{L}=3000~(300)\,{\rm fb}^{-1}\, ,
\ee
Therefore, a discovery of Higgs pair production at the
HL-LHC might be possible if the
mis-identification of light and charm jets as
$b$-jets can be  reduced.
%
With these assumptions, and provided
the signal efficiency can be further enhanced,
this measurement might even be possible
by the end of Run II.
%
The smaller value of $\la n_{\rm PU}\ra$ at Run II as compared
to the HL-LHC enviroment should be beneficial in this
respect.
